---
title: "Neighborhood in MAC"
author: 
date: "September, 15th 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, load data, warning=F, message=F, echo=F}



library(likelihood)
library(forcats)
library(dplyr)
library(reshape2)
library(knitr)
library(ggplot2)
library(kableExtra)


load ("../Data/NCI_MAC.Rdata")
```


We want to characterize how is the neighborhood of each species. More precisely, we want to define 

1) How many trees of each species are in the neighborhood of a certain target species
2) How the functional diversity varies among target species


### (1) Specific neighborhood

We could first calculate how many neighbours has a tree of a certain species on average, and compare the results across species

```{r average neighbours, fig.width=10}

ggplot(targets) +
        geom_boxplot( aes (CodeSp, n_neighbours)) 
        
```

 We can see there are no big differences, as expected. We can now check how many neighbours of each species has a given target species
 
```{r species neighborhood, size = 'tiny'}
neighbour_sps <- table (neighbours$CodeSp_IN, neighbours$CodeSp_NEAR)
kable(addmargins(neighbour_sps,1), format ="html") %>%
        kable_styling(bootstrap_options =  c("striped","hover", "condensed", "responsive"), full_width = F, font_size = 12,
                      position = "left") %>%
        column_spec(1, bold = T, border_right = T)
```

In this case, we see that some species are indeed very rare. Since our most complex model has 30 parameters, and some studies recommend that the ratio observations/parameters should be > 10, I decided to collapse all neighbouring species with less than 300 observations into a "other" group.


### (2) Diversity neighborhood

It is time now to assess if the functional diversity neighborhood is very different across target species. Remember we defined species richness using Shannon and two FD indicators: functional difference and functional dispersion. Let's see each of them

```{r sps richness, fig.width=10, warning=F, message=FALSE}

ggplot(targets) +
        geom_boxplot( aes (CodeSp, shannon),outlier.color = NA) +
        ylim(c(0,10)) +
        ggtitle ("Species richness (Shannon)")
```


```{r func diff, fig.width=10}

ggplot(targets) +
        geom_boxplot( aes (CodeSp, FDif2), outlier.color = NA) +
        ggtitle ("Functional difference")
```

```{r func dispersion, fig.width=10}

ggplot(targets) +
        geom_boxplot( aes (CodeSp, FDis), outlier.colour = NA) +
        ggtitle ("Functional dispersion")
``` 


In general, I believe that, despite differences across species, values are similar enough as to don't correct FD values.




